Location and Scale-Invariant Power Transformations for Transforming Data to

### 2025-10-24 manuscript_rmarkdown/manuscript.Rmd
Rework section 2.4.

---
title: "Location and Scale-Invariant Power Transformations for Transforming Data to
  Normality"
author: "Alex Zwanenburg, Steffen Löck"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: lualatex
    keep_tex: true
papersize: a4
bibliography: refs.bib
header-includes:
- \usepackage{amsmath}
- \usepackage{amsthm}
- \usepackage{booktabs}
- \newtheorem*{definition}{Definition}
- \DeclareMathOperator*{\argmax}{argmax}
- \DeclareMathOperator*{\argmin}{argmin}
- \DeclareMathOperator{\sgn}{sgn}
---

@@

## Central normality test

Power transformations aim to transform features to a normal distribution.
However, this may not always be successful or possible.
Deviations from normality can be detected by normality tests, such as the Shapiro-Wilk test [@Shapiro1965-zd].
In practice, normality tests may be too stringent with large sample sizes, outliers, or both.
Here we develop a test for central normality.

@@

\begin{definition}[Central normality of a sequence]

The central portion of a sequence is normally distributed if the sum of residual
errors is equal to zero:
$\sum_{r_j \in \mathbf{R}_{\text{c}}} r_j = 0$

@@

In practice, a finite sequence sampled from a normal distribution $\mathcal{N}(\mu, \sigma)$ will have a non-zero sum of residual errors.

\begin{definition}[Central normality test]
The null-hypothesis is defined as: $\mathcal{H}_0: \sum_{r_j \in \mathbf{R}_{\text{c}}}r_j = 0$

The alternative hypothesis is defined as: $\mathcal{H}_1: \sum_{r_j \in \mathbf{R}_{\text{c}}}r_j > 0$
\end{definition}

The null-hypothesis $\mathcal{H}_0$ is that the central portion of sequence $\mathbf{X}$
is normally distributed, with the alternative hypothesis $\mathcal{H}_1$ that it
is not normally distributed. To test this hypothesis, a test statistic is computed
and compared against a critical value.

\begin{definition}[Central normality test statistic]
Let $m$ be the number of elements of $\mathbf{R}_{\text{c}}$

@@

The null-hypothesis should be rejected at significance level $\alpha$ if 
$\tau_{n, \kappa} \geq \tau_{\alpha, n, \kappa, \text{critical}}$.

The central portion of the data needs to be defined and the Type 1 error rates determined to provide critical test statistics.
We will do so in the [Simulation](#simulation) section.

@@

The critical test statistic for the empirical central normality test depends on 
the significance level $\alpha$, sequence length $n$, central portion $\kappa$
and estimated outlier rate $\zeta$, and was set through simulation. 
For each number of samples $n \in \left\{\lfloor 10^\nu \rfloor | \nu \in \left\{0.7500, 0.8125, \ldots, 4.0000 \right \} \right\}$
we randomly sampled $m_d = 30000$ ordered sequences $\mathbf{X}$ from $\mathcal{N}(0,1)$.
Outliers were inserted by randomly replacing elements of an ordered sequence with probability 
$\zeta \in \left\{0.0, 0.01, 0.02, 0.05, 0.10, 0.15, 0.20 \right\}$. 
Outlier values were set according to @Tukey1977-xm, as follows.
Let $x^{*} \sim U\left(-2, 2\right)$.
Then the corresponding outlier value was:

>>>>

---
title: "Location and Scale-Invariant Power Transformations for Transforming Data to
  Normality"
author: "Alex Zwanenburg, Steffen Löck"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: lualatex
    keep_tex: true
papersize: a4
bibliography: refs.bib
header-includes:
- \usepackage{amsmath}
- \usepackage{amsthm}
- \usepackage{booktabs}
- \newtheorem*{definition}{Definition}
- \newtheorem*{corollary}[Corollary]
- \DeclareMathOperator*{\argmax}{argmax}
- \DeclareMathOperator*{\argmin}{argmin}
- \DeclareMathOperator{\sgn}{sgn}
---

@@

Power transformations aim to transform features to a normal distribution.
However, this may not always be successful or possible.
Deviations from normality can be detected by normality tests, such as the Shapiro-Wilk test [@Shapiro1965-zd].
In practice, normality tests may be too stringent with large sample sizes, outliers, or both.
Here we develop an empirical test for central normality.

@@

\begin{definition}[Centrally normal distribution]

A centrally normal distribution is any distribution whose quantile function is 
defined by $F^{-1}_{\mathcal{N}}$ for $\frac{1-\kappa}{2} \leq  p \leq \frac{1 + \kappa}{2}$,
with $\mathcal{N}$ any normal distribution.

@@

\begin{corollary}

If the sum of residual errors of the central portion of a sequence is 0, i.e.,
$\sum_{r_j \in \mathbf{R}_{\text{c}}} r_j = 0$, 
the values of the elements in the central portion of that sequence follow a 
centrally normal distribution.

\end{corollary}

By the Glivenko-Cantelli theorem, if sequence $\mathbf{X}$ was sampled from a
normal distribution $\mathcal{N}(\mu, \sigma)$,
$\lim_{n \rightarrow \infty} \sum_{r_j \in \mathbf{R}_{\text{c}}} r_j = 0$,
i.e. an infinite sequence sampled from a normal distribution has a central portion
with values sampled from a centrally normal distribution.

\begin{definition}[Central normality test]
The null-hypothesis $\mathcal{H}_0$ is that the sequence $\mathbf{X}$ was sampled
from a population that has a centrally normal distribution.

The alternative hypothesis is that the sequence $\mathbf{X}$ was sampled from a
population that lacks a centrally normal distribution.
\end{definition}

To test this hypothesis, a test statistic is computed and compared against a critical value.

\begin{definition}[Central normality test statistic]
Let $m$ be the number of elements of $\mathbf{R}_{\text{c}}$.

@@

The null-hypothesis should be rejected at significance level $\alpha$ if 
$\tau_{n, \kappa} \geq \tau_{\alpha, n, \kappa, \zeta, \text{critical}}$,
with $\zeta$ the population estimate of the rate of outliers. Critical test
values are determined in the [Simulation](#simulation) section.

@@

The critical test statistic for the empirical central normality test depends on 
the significance level $\alpha$, sequence length $n$, central portion $\kappa$
and estimated outlier rate $\zeta$.
Critical test statistic values were set through simulation, as described below.

For each number of samples $n \in \left\{\lfloor 10^\nu \rfloor | \nu \in \left\{0.7500, 0.8125, \ldots, 4.0000 \right \} \right\}$
we randomly sampled $m_d = 30000$ ordered sequences $\mathbf{X}$ from $\mathcal{N}(0,1)$.
Outliers were inserted by randomly replacing elements of an ordered sequence with probability 
$\zeta \in \left\{0.0, 0.01, 0.02, 0.05, 0.10, 0.15, 0.20 \right\}$. 
Outlier values were set according to @Tukey1977-xm, as follows.
Let $x^{*} \sim U\left(-2, 2\right)$.
Then the corresponding outlier value was:
